---
title: "Enrichment NB 02: Using rSEA to analyze pathway enrichment for CPTAC CNV data"
author: "Caleb Lindgren"
output:
  pdf_document: default
---
# \*\*\*WARNING\*\*\*: Bad analysis technique

I realized that this folder of analyses falls prey to the fallacy of indirect comparison. We compare with_event_tumor to normal, and without_event_tumor to normal, and then compare the with_tumor and without_tumor results from that to each other. It's better to just directly compare with_tumor to without_tumor, which we do with a t-test in the finished analyses.

# Enrichment NB 02: Using rSEA to analyze pathway enrichment for CPTAC CNV data

## Instructions for running:

The best way to run this R Markdown notebook is using the RStudio IDE. Start RStudio, open this file, and then press the green play buttons in the top right corners of the code cells. Alternatively, you can run it from the R command line interface using the `rmarkdown` package.

## Setup

```{r, message=FALSE}
library(tidyverse)
library(rSEA)

THRESHOLD <- 0.99 # Since we've filtered our feature sets to include only truly active ones; but there's a bug such that
# a threshold of 1 causes an exception

INPUT_DIR <- "results_01_grouped_expr"
INPUT_FILE <- "unique_results.tsv.gz"
INPUT_FILE_PATH <- file.path(INPUT_DIR, INPUT_FILE)

GENE_SETS_DIR <- "gene_set_libraries"
GENE_SET_FILES <- c("GO_Biological_Process_2018.gmt", "ReactomePathways.gmt")

OUTPUT_DIR <- "results_02_rSEA"

# Create the output directory if it doesn't already exist
dir.create(OUTPUT_DIR, showWarnings = FALSE)
```

## Step 1: Load expression data and divide by cancer type and group

```{r, message=FALSE, warning=FALSE}
all_expr_data <- read_tsv(INPUT_FILE_PATH)
groups <- distinct(all_expr_data, cancer_type_group)$cancer_type_group
split_expr_data <- lapply(groups, function(x) filter(all_expr_data, cancer_type_group == x))
names(split_expr_data) <- groups
```

## Step 2: Load pathway lists

GO downloaded on 25 Jul 2020 from: https://amp.pharm.mssm.edu/Enrichr/geneSetLibrary?mode=text&libraryName=GO_Biological_Process_2018
Reactome downloaded on 27 Jul 2020 from: https://reactome.org/download/current/ReactomePathways.gmt.zip

Each line of the Reactome file was reformatted to have the pathway ID and name both in the first field, using the find regex `"^([^\t]*)\t(R-HSA-\d+)\t"` and then the replace string `"\1 (\2)\t\t"`.

## Step 3: Run rSEA and save the results

For some reason, our data was generating an error when the threshold was set to 1. Some problem in the `hommel` package. Since the threshold is set to 1 whenever the competitive hypothesis is tested, we set the optional `competitive` parameter to `FALSE`, to override the default behavior of automatically testing the competitive hypothesis.

```{r}
for (gene_set_file in GENE_SET_FILES) {
  
  # Step 2
  gene_set_file_path <- file.path(GENE_SETS_DIR, gene_set_file)
  
  pathways <- gene_set_file_path %>%
    readLines() %>% # Read in each line of the file as a string
    sapply(strsplit, split = "\t") # Split each line on the tab character
  
  pathway_ids <- sapply(pathways, function(x) x[[1]]) # Save the pathway name and id field for later
  pathways <- sapply(pathways, function(x) x[-(1:2)]) # Cut out the pathway names and blank field from the lists
  names(pathways) <- pathway_ids # Set the pathway ids as the names
  
  # Step 3
  results <- lapply(split_expr_data, function(x) SEA(
    pvalue = x$adj_p,
    featureIDs = x$protein_str,
    pathlist = pathways,
    thresh = THRESHOLD,
    competitive = FALSE))
  
  names(results) <- groups
  
  results <- groups %>%
    lapply(function(x) results[[x]] %>% add_column(cancer_type_group = x)) %>% # Add each table's cancer type group as a column
    bind_rows() # Combine all the tables into one table
  
  OUTPUT_FILE_PATH <- file.path(
    OUTPUT_DIR,
    paste("enrichment_rsea_thresh_", THRESHOLD, "_lib_", strsplit(gene_set_file, "\\.")[[1]][[1]], ".tsv", sep = ""))
  
  write_tsv(results, OUTPUT_FILE_PATH)
}
```